Exploring the conformational dynamics of alanine dipeptide in solution subjected to an external electric field: A nonequilibrium molecular dynamics simulation

### 2014-03-05 writedown/npt-neq/alanine.tex
2nd round

\section{Introduction}
The possible effect of the electric field (EF) on the conformation of
proteins is a subject of increasing attention since it is linked to the problem of exposing human tissues to
electromagnetic radiation. Despite its major relevance, a
satisfactory understanding of how the EF influences the behaviour of proteins
has not been reached so far although a considerable number of
experimental e.g.~\cite{bohr2000microwave, bohr2000microwave-1,
  dePomerai2000cell, inskip2001cellular, mancinelli2004non} and
theoretical e.g.~\cite{budi2005electric, budi2007effect,
  budi2008comparative, toschi2008effects, astrakas2011electric,
  astrakas2012structural, damm2012can, starzyk2013proteins,
  english2009nonequilibrium, solomentsev2012effects}
studies have been devoted to this topic.
In this perspective, molecular dynamics (MD) has been proven useful in
understanding the behaviour of a protein 
in both static and oscillatory EF~\cite{budi2005electric, budi2007effect, budi2008comparative,
  toschi2008effects, astrakas2011electric, astrakas2012structural,
  damm2012can, starzyk2013proteins, english2009nonequilibrium,
  solomentsev2012effects}. In these studies, the initial
configurations are usually taken from the Protein Data Bank (PDB),
and then temperature controlled simulations are performed for a maximum
time range of the order of 10-60 nanoseconds.
During this simulation time 
relevant conformational changes characterizing the solvated molecule can be observed. 
Along the trajectories, various observable, such as the
secondary structure, root mean square displacement (RMSD), dipole
moment, shape parameters are calculated as a function of time, or
in stationary condition, taking time averages.
This approach releases useful information on how the
secondary structure of the molecule is changed by external  EF.  However,
the behaviour of observables in time is characterized by 
strong
fluctuations. The entity of the fluctuations is such that a quantitative 
analysis becomes difficult and does not allow for  a clear understanding of the rationale behind the conformational changes.
A time average can reduce the statistical uncertainty,
however in non-stationary conditions this would be a not-well-founded procedure.
In fact, we are studying non-stationary nonequilibrium process
(e.g. relaxing process to the new conformations under a static
EF, or the dynamics from one conformation to the other under an oscillatory EF).
In this perspective, if one wants to use MD, the proper way to do it would be that of treating a situation of nonequilibrium with algorithms that can explicitly describe it.
This is the approach that will be followed in this work, using
the dynamical nonequilibrium molecular dynamics simulation (D-NEMD).
The method, firstly developed by one of us and coworkers~\cite{ciccotti1975direct, ciccotti1979thought, ciccotti1993theoretical, palla2008bulk}, 
has been recently applied, to the study of
hydrodynamics~\cite{orlandini2011hydrodynamics,
  orlandini2011hydrodynamics-01}. The essential feature of D-NEMD is to
provide a way of evaluating time-dependent
nonequilibrium averages starting from a properly determined initial ensemble.
In this context, the method can be used to analyze the
conformational changes of a solvated molecule, subject to a time-dependent EF.

@@

\section{Methodology}
\subsection{Nonequilibrium molecular dynamics simulation}
In this section we briefly review the dynamical approach to nonequilibrium
molecular dynamics (D-NEMD) \cite{ciccotti1975direct, ciccotti1979thought,
  orlandini2011hydrodynamics, orlandini2011hydrodynamics-01}.
In the following we denote the macroscopic observable by $O(t)$. If at $t$
the configurational probability distribution is $\rho(\vect x, t)$, where
$\vect x$ is the phase space variable, then the observable can be
expressed by
\begin{align}\label{eqn:tmp1}
  O(t) = \int d\vect x\, \hat O(\vect x)\rho(\vect x, t)  = \langle \hat O(\vect x), \rho(\vect x, t)\rangle,
\end{align}
we refer to $\hat O (\vect x)$ as the microscopic observable, 
which is measured at each point in the phase space $\vect x$.
Here we always assume that the initial probability distribution
$\rho(\vect x, 0)$ is known. In particular, in our case,
it is identical to the equilibrium distribution of the system
without EF.
The bracket on the right hand side of Eq.~\eqref{eqn:tmp1} denotes, as usual, the inner product in the
phase space.  We assume that the dynamics of the system is governed by the
Hamiltonian equation, i.e. $\dot {\vect x} = J \cdot \nabla_{\vect x}
\mathcal H(\vect x, t)$, where $\mathcal H$ is the time-dependent Hamiltonian,
and $J$ is
the symplectic matrix. The Liouville equation for the probability
distribution writes:
\begin{align}\label{eqn:tmp2}
  \frac{\partial \rho(\vect x, t)}{\partial t} = - iL(t) \rho(\vect x, t),
\end{align}
where $iL(t) = \{\cdot, \mathcal H\}$ is the Liouville operator.
Eq.~\eqref{eqn:tmp2}
can be formally solved by $\rho(\vect x, t) = U^\dagger(t,0) \rho(\vect x, 0)$,
{where $U^\dagger(t,0) = \mathcal T\exp\{-i \int_0^t d t' L(t')\}$, and $\mathcal T$ is the time ordering operator.}
On the other hand,
\begin{align}
  \frac{d \hat O(\vect x(t))}{dt} = \nabla_{\vect x}\hat O\cdot \dot{\vect x}
  = \nabla_{\vect x}\hat O\cdot J\cdot \nabla_{\vect x}\mathcal H
  = iL(t) \hat O (\vect x(t)).
\end{align}
This equation can be formally solved by
$\hat O(\vect x(t)) = U(t,0)\,\hat O(\vect x(0))$, therefore,
\begin{align}\nonumber
  O(t) & = \langle \hat O(\vect x), \rho(\vect x, t)\rangle
  = \langle \hat O(\vect x), U^\dagger(t,0)\, \rho(\vect x, 0)\rangle
  = \langle U(t,0)\,\hat O(\vect x), \rho(\vect x, 0)\rangle\\\label{eqn:tmp4}
  &= \langle \hat O(\vect x(t)), \rho(\vect x, 0)\rangle.
\end{align}
Since we assume that the system starts from the equilibrium distribution (without EF)
 Eq.~\eqref{eqn:tmp4} expresses the fact that the 
observable $O(t)$ calculated in a situation of nonequilibrium is equal to the ensemble average of
the microscopic observable computed along trajectories, starting from
initial configurations sampled from an \emph{equilibrium}
distribution. In practical terms we proceed by first running an equilibrium MD simulation
to generate a sample of configurations. Next we employ these configurations as
initial configurations and, for each, the full Hamiltonian dynamics is integrated until
time $t$.
In this paper, these trajectories are called \emph{branching
  trajectories}.
% {\bf EITHER WE PUT THE PICTURE OR REMOVE THIS SENTENCE {(see Fig.\ref{branching} for a pictorial representation)}}. 
Finally, the value of the macroscopic observable at time $t$ is
calculated by averaging the microscopic observable measured at each time $t$ along each branching trajectory.
Stochastic dynamics (e.g.~Langevin dynamics) can be handled analogously.

@@

\subsection{System settings and simulation protocol}
The system is set up in a $2.7\times 2.7\times
2.7\, \textsf{nm}^3$ periodic simulation region, with one alanine dipeptide
described by the CHARMM27 force field~\cite{foloppe2000all}, and dissolved in 644 TIP3P~\cite{jorgensen1983comparison}
water molecules.  \recheck{The grid-based energy correction map (CMAP)~\cite{mackerell2004extending}
for the backbone dihedral angles is also used.}
%The EF exserts force $\vect F = q \vect E(t)$ on
%the partial charge $q$ in the system.
The size of the  dynamical region is $R_{ex} = 1.0$~nm.
All simulations are performed by a home-modified GROMACS~4.5~\cite{pronk2013gromacs} with support to the CHARMM
force field~\cite{bjelkmar2010implementation}.
First, an equilibrium NVT simulation at 300~K of
100~\textsf{ns} was performed with a Langevin thermostat (time-scale
$\tau_T = 0.5~\textsf{ps}$).  Along the trajectory, configurations were taken every
50~\textsf{ps} and we used 2000 initial configurations for each nonequilibrium
MD simulation (if not stated otherwise for specific cases).
The branching trajectories were integrated by the
Leap-frog scheme with the aforementioned nonequilibrium
temperature control technique.  The
time step was $0.002~\textsf{ps}$.
\recheck{The short-range
interaction (van der Waals interaction)
have a cut-off radius of 1.0~nm, and has been smoothed from $0.8$ to $1.0$~nm by the
``switch'' method provided by the GROMACS code.
A energy conserving particle mesh Ewald method (PME)~\cite{darden1993pme, essmann1995spm}
method was applied to calculate the electrostatic interaction in this
periodic system. For the direct space part of PME the cut-off and smoothing follow the same principles as those applied to the van der Waals interaction.}
\recheck {All hydrogen involving bonds are constrained by LINCS~\cite{hess1997lincs}, except 
  the TIP3P water molecules which are constrained by the SETTLE algorithm~\cite{miyamoto2004settle}.}
In the thermostated region, the original dynamics was
coupled to a Langevin thermostat with $\tau_T = 0.1~\textsf{ps}$.
  In all testing cases, this local thermostat is able to control
  the system at the desired temperature, i.e.~300~K (the results
  are not presented in this paper).
The whole system is also coupled to a Parrinello-Rahman barostat~\cite{parrinello1981polymorphic} with $\tau_P = 2.0~\textsf{ps}$ to keep
the pressure at ambient condition (1~Bar). Since the
change of the system size is small and slow, the pressure control
does not have an sizable effect on the
dynamics of the system.
At time $t=0$~ps, the system
has been fully equilibrated without any EF. From $t=0$ to
the warm-up time $t=t_{\textrm{warm}}$, the EF is switched on linearly, while
after $t=t_{\textrm{warm}}$, the field is kept constant in time at
$E_{\infty}$.  In this work we consider $t_{\textrm{warm}} = 10$~ps,
and $E_{\infty} = 1$~V/nm.
The direction of the field is arbitrarily chosen along the
$x$ direction.
In this paper we denote the
EF as a function of time $\vect E(t)$.
Therefore, for the case of constant EF, we have
$\vect E(t) = (E_\infty\cdot t/t_{warm},0,0)$ for $0\leq t < t_{warm}$, and 
$\vect E(t) = (E_\infty,0,0)$
for $t \geq t_{warm}$.

@@

The change of conformation in time of the alanine dipeptide is
investigated by analyzing the probability density $p(\phi,\psi,t)$ on
the Ramachandran plot at time $t$. The definition of the dihedral
angles $\phi$ and $\psi$ is given in Fig.~\ref{fig:tmp4}~(a).
The probability density of conformations of equilibrium in absence of external field is shown in Fig.~\ref{fig:tmp4}~(b), while the same quantity, resulting from the molecular relaxation to the action of $E_{\infty} = 1$~V/nm is
given in Fig.~\ref{fig:tmp4}~(c). In the plot one can identify several clusters
in which conformations are grouped.
Therefore, in order to simplify the analysis, we divide the
Ramachandran plot into 5 subregions $\{\confaa, \confab, \confba, \confbb, \confc\}$
(see caption of Fig.~\ref{fig:tmp4} for the corresponding definition).
It is worth to notice that these conformations
are found in different secondary structures of a peptide chain:
$\confaa$ and $\confab$ are
corresponding to the $\alpha$-helix. $\confba$ and $\confbb$ are
corresponding to the $\beta$-sheet. $\confc$ is corresponding to the
left-handed $\alpha$-helix.

From Fig.~\ref{fig:tmp4}, it is evident that the probability with which a
conformation occurs changes when the external EF is applied.
The most evident case is conformation $\confc$; in fact it is {almost} not present before the external field is applied but appears in a clear way as a response to the action of the EF. Following this line of thought, we are interested to study the change of conformation under the action of EF. Specifically, for each conformational change, we will analyze the relation between the structural relaxation of the molecule and its corresponding time scale. 
The starting point is the calculation of the probability
of each conformation:
\begin{align}
  P_I(t) = \mathbb P ( (\phi_t,\psi_t) \in I), \quad  I \in \{\confaa, \confab, \confba, \confbb, \confc\},
\end{align}
Notice that the probability is time dependent, being an observable in nonequilibrium situation.
Next,
we consider the net probability flux from conformation $I$ to $J$, defined by:
\begin{align}\nonumber
  F_{J,I}(t) & = \frac1{\Delta t}\,
  \bigg[\,
  \mathbb P \big( (\phi_{t-\Delta t},\psi_{t-\Delta t}) \in J,\, (\phi_t, \psi_t) \in I \,\big) 
  -
  \mathbb P \big( (\phi_{t-\Delta t}, \psi_{t-\Delta t}) \in I,\, (\phi_t, \psi_t) \in J \,\big)
  \bigg], \\\label{eqn:tmp7}
  & J,I \in \{\confaa, \confab, \confba, \confbb, \confc\}.
\end{align}
% {\bf Here for $\lim_{\Delta \to 0}$ is intended that in the simulation the time interval is taken large enough so that we have  enough events and thus stable numbers (see also Ref.~\cite{schuette2011markov})}.
Where $\mathbb P\big( (\phi_{t-\Delta t}, \psi_{t-\Delta t}) \in J,\, (\phi_t, \psi_t) \in I\,\big)$
is the joint probability of the alanine being in conformation $J$ at
time $(t-\Delta t)$ \emph{and} being in conformation $I$ at time $t$.
The same definition applies to
$\mathbb P \big( (\phi_{t-\Delta t}, \psi_{t-\Delta t}) \in I,\, (\phi_t, \psi_t) \in J\,\big)$.
Here the time interval $\Delta t$ should be small enough compared to the
time scale of conformational dynamics,
so that the changes in the conformation probabilities
can be treated in linear approximation.
At the same time,  $\Delta t$ should be also large enough compared with
the time step of the MD integrator, 
so that the quantities of Eq.~\eqref{eqn:tmp7} are well-defined, and can be
estimated with sufficient numerical accuracy (see also Ref.~\cite{schuette2011markov}).
\recheck{In all the numerical examples of this work, $\Delta t = 1$~ps is used}.
A positive value of $F_{J,I}(t)$ indicates that the
net flux goes from $J$ to $I$,
while a negative value indicates a net flux from $I$ to $J$.
Through $F_{J,I}(t)$, the analysis of the nonequilibrium process 
is projected onto the analysis of the probabilistic link between discretized conformations $\{\confaa, \confab, \confba, \confbb, \confc\}$.
\recheck{ The concept of net probability flux employed by us is similar to the
  concept of kinetic rate.  However, it must be underlined that, in general, the
  probability flux should not be interpreted as a kinetic rate.
  Nonequilibrium molecular dynamics cannot be replaced by 
  kinetic analysis, the main reason is, firstly, that the nonequilibrium
  nature of the problem requires the use of a small time interval $\Delta t$ so that the probability flux is well defined.
  Secondly, within the small
  time interval (we use 1~ps), the conformational
  dynamics of alanine dipeptide is generally not Markovian~\cite{schuette2011markov, prinz2011markov},
  therefore, a direct application of the kinetic
  analysis may be questionable.}
In the case of oscillatory EF, the behaviour of the net probability flux is also highly oscillatory, thus, in order to clearly identify the essential features of the molecular conformational changes, it is convenient to study the integrated net probability flux, defined as:
\begin{align}\label{eqn:tmp7a}
  Q_{J,I} (t) = \int_0^t F_{J,I}(\tau)d \tau,
\end{align}
which expresses the cumulative effects due to the action of EF on the conformations of the molecule.

@@

\begin{figure}
  \centering
  \includegraphics[]{fig.ala.new/fig-meta-more.eps}
  \caption{The time-dependent probability of conformations $P_I(t)$ 
    when a constant EF is present, where $\sum_I P_I(t) = 1, \forall t$.
    The warm-up time is $t_{warm} = 10$~ps. The red line refers to conformation $I = \confaa$,
    green to $\confab$, dark blue to $\confba$, purple to $\confbb$ and light blue
    to $\confc$. \recheck{
      The solid lines denote the results of one branching trajectory
      starting from each initial conformation, while the dashed lines denote
      the results of four branching trajectories (different random seeds for the Langevin thermostat) starting from each initial
      conformation.
      The shadowed region with each solid line denotes the
      statistical uncertainty of the result at 95\% confidence level.
    }
  }
  \label{fig:tmp5}
\end{figure}

>>>>

\section{Introduction}
The possible effect of the electric field (EF) on the conformation of
proteins is a subject of increasing attention since it is linked to the problem of exposing human tissues to
electromagnetic radiation. Despite its major relevance, a
satisfactory understanding of how the EF influences the behaviour of proteins
has not been reached so far although a considerable number of
experimental e.g.~\cite{bohr2000microwave, bohr2000microwave-1,
  dePomerai2000cell, inskip2001cellular, mancinelli2004non} and
theoretical e.g.~\cite{budi2005electric, budi2007effect,
  budi2008comparative, toschi2008effects, astrakas2011electric,
  astrakas2012structural, damm2012can, starzyk2013proteins,
  english2009nonequilibrium, solomentsev2012effects}
studies have been devoted to this topic.
In this perspective, molecular dynamics (MD) has been proven useful in
understanding the behaviour of a protein 
in both static and oscillatory EF~\cite{budi2005electric, budi2007effect, budi2008comparative,
  toschi2008effects, astrakas2011electric, astrakas2012structural,
  damm2012can, starzyk2013proteins, english2009nonequilibrium,
  solomentsev2012effects}. In these studies, the initial
configurations are usually taken from the Protein Data Bank (PDB),
and then temperature controlled simulations are performed for a maximum
time range of the order of 10-60 nanoseconds.
During these simulations time 
relevant conformational changes characterizing the solvated molecule can be observed. 
Along the trajectories, various observable, such as the
secondary structure, root mean square displacement (RMSD), dipole
moment, and shape parameters are calculated as a function of time, or
in stationary condition, taking time averages.
This approach releases useful information on how the
secondary structure of the molecule is changed by external  EF.  However,
the behaviour of observables in time is characterized by 
strong
fluctuations. The entity of the fluctuations is such that a quantitative 
analysis becomes difficult and does not allow for  a clear understanding of the rationale behind the conformational changes.
A time average can reduce the statistical uncertainty,
however in non-stationary conditions this would be a not-well-founded procedure.
In fact, we are studying non-stationary nonequilibrium process
(e.g. relaxing process to the new conformations under a static
EF, or the dynamics from one conformation to the other under an oscillatory EF).
In this perspective, if one wants to use MD, the proper way to do it would be that of treating a situation of nonequilibrium with algorithms that can explicitly describe it.
This is the approach that will be followed in this work, using
the dynamical nonequilibrium molecular dynamics simulation (D-NEMD).
The method, firstly developed by one of us and coworkers~\cite{ciccotti1975direct, ciccotti1979thought, ciccotti1993theoretical, palla2008bulk}, 
has been recently applied, to the study of
hydrodynamics~\cite{orlandini2011hydrodynamics,
  orlandini2011hydrodynamics-01}. The essential feature of D-NEMD is to
provide a way of evaluating time-dependent
nonequilibrium averages starting from a properly determined initial ensemble.
In this context, the method can be used to analyze the
conformational changes of a solvated molecule, subject to a time-dependent EF.

@@

\section{Methodology}
\subsection{Dynamical nonequilibrium molecular dynamics simulation (D-NEMD)}
In this section we briefly review the dynamical approach to nonequilibrium
molecular dynamics (D-NEMD) \cite{ciccotti1975direct, ciccotti1979thought,
  orlandini2011hydrodynamics, orlandini2011hydrodynamics-01}.
In the following we denote the macroscopic observable by $O(t)$. If at $t$
the configurational probability distribution is $\rho(\vect x, t)$, where
$\vect x$ is the phase space variable, then the observable can be
expressed by
\begin{align}\label{eqn:tmp1}
  O(t) = \int d\vect x\, \hat O(\vect x)\rho(\vect x, t)  = \langle \hat O(\vect x), \rho(\vect x, t)\rangle,
\end{align}
we refer to $\hat O (\vect x)$ as the microscopic observable, 
which is measured at each point in the phase space $\vect x$.
Here we always assume that the initial probability distribution
$\rho(\vect x, 0)$ is known. In particular, in our case,
it is identical to the equilibrium distribution of the system
without EF.
The bracket on the right hand side of Eq.~\eqref{eqn:tmp1} denotes, as usual, the inner product in the
phase space.  We assume that the dynamics of the system is governed by the
Hamiltonian equation, i.e. $\dot {\vect x} = J \cdot \nabla_{\vect x}
\mathcal H(\vect x, t)$, where $\mathcal H$ is the time-dependent Hamiltonian,
and $J$ is
the symplectic matrix
\begin{align}
  J = \left(
    \begin{array}{rr}
      0 & I\\
      -I & 0
    \end{array}
    \right).
\end{align}
The Liouville equation for the probability
distribution writes:
\begin{align}\label{eqn:tmp2}
  \frac{\partial \rho(\vect x, t)}{\partial t} = - iL(t) \rho(\vect x, t),
\end{align}
where $iL(t) = \{\cdot, \mathcal H\}$ is the Liouville operator.
Eq.~\eqref{eqn:tmp2}
can be formally solved by $\rho(\vect x, t) = U^\dagger(t,0) \rho(\vect x, 0)$,
{where $U^\dagger(t,0) = \mathcal T\exp\{-i \int_0^t d t' L(t')\}$, and $\mathcal T$ is the time ordering operator.}
On the other hand,
\begin{align}
  \frac{d \hat O(\vect x(t))}{dt} = \nabla_{\vect x}\hat O\cdot \dot{\vect x}
  = \nabla_{\vect x}\hat O\cdot J\cdot \nabla_{\vect x}\mathcal H
  = iL(t) \hat O (\vect x(t)).
\end{align}
This equation can be formally solved by
$\hat O(\vect x(t)) = U(t,0)\,\hat O(\vect x(0))$, therefore,
\begin{align}\nonumber
  O(t) & = \langle \hat O(\vect x), \rho(\vect x, t)\rangle
  = \langle \hat O(\vect x), U^\dagger(t,0)\, \rho(\vect x, 0)\rangle
  = \langle U(t,0)\,\hat O(\vect x), \rho(\vect x, 0)\rangle\\\label{eqn:tmp4}
  &= \langle \hat O(\vect x(t)), \rho(\vect x, 0)\rangle.
\end{align}
Since we assume that the system starts from the equilibrium distribution (without EF)
 Eq.~\eqref{eqn:tmp4} expresses the fact that the 
observable $O(t)$ calculated in a situation of nonequilibrium is equal to the ensemble average of
the microscopic observable computed along trajectories, starting from
initial configurations sampled from an initial or, sometimes, an \emph{equilibrium}
distribution. In practical terms we proceed by first running an equilibrium MD simulation
to generate a sample of configurations. Next we employ these configurations as
initial configurations and, for each, the full Hamiltonian dynamics is integrated until
time $t$.
In this paper, these trajectories are called \emph{branching
  trajectories}.
% {\bf EITHER WE PUT THE PICTURE OR REMOVE THIS SENTENCE {(see Fig.\ref{branching} for a pictorial representation)}}. 
Finally, the value of the macroscopic observable at time $t$ is
calculated by averaging the microscopic observable measured at each time $t$ along each branching trajectory.
Stochastic dynamics (e.g.~Langevin dynamics) can be handled analogously.

@@

\subsection{System settings and simulation protocol}
The system is set up in a $2.7\times 2.7\times
2.7\, \textsf{nm}^3$ periodic simulation region, with one alanine dipeptide
described by the CHARMM27 force field~\cite{foloppe2000all}, and dissolved in 644 TIP3P~\cite{jorgensen1983comparison}
water molecules.  \recheck{The grid-based energy correction map (CMAP)~\cite{mackerell2004extending}
for the backbone dihedral angles is also used.}
%The EF exserts force $\vect F = q \vect E(t)$ on
%the partial charge $q$ in the system.
The size of the  dynamical region is $R_{ex} = 1.0$~nm.
All simulations are performed by a home-modified GROMACS~4.5~\cite{pronk2013gromacs} together with the CHARMM
force field~\cite{bjelkmar2010implementation}.
First, an equilibrium NVT simulation at 300~K of
100~\textsf{ns} was performed with a Langevin thermostat (time-scale
$\tau_T = 0.5~\textsf{ps}$).  Along the trajectory, configurations were taken every
50~\textsf{ps} and we used 2000 initial configurations for each nonequilibrium
MD simulation (if not stated otherwise for specific cases).
The branching trajectories were integrated by the
Leap-frog scheme \bluec{(standard Gromacs implementation)} with the aforementioned nonequilibrium
temperature control technique.  The
time step was $0.002~\textsf{ps}$.
\recheck{The short-range
interaction (van der Waals interaction)
have a cut-off radius of 1.0~nm, and has been smoothed from $0.8$ to $1.0$~nm by the
``switch'' method provided by the GROMACS code.
A energy conserving Particle Mesh Ewald (PME)~\cite{darden1993pme, essmann1995spm}
method was applied to calculate the electrostatic interaction in this
periodic system. For the direct space part of PME the cut-off and smoothing follow the same principles as those applied to the van der Waals interaction.}
\recheck {All hydrogen involving bonds are constrained by LINCS~\cite{hess1997lincs}, except 
  the TIP3P water molecules which are constrained by the SETTLE algorithm~\cite{miyamoto2004settle}.}
In the thermostated region, the original dynamics was
coupled to a Langevin thermostat with $\tau_T = 0.1~\textsf{ps}$.
  In all testing cases, this local thermostat is able to control
  the system at the desired temperature, i.e.~300~K (the results
  are not presented in this paper).
The whole system is also coupled to a Parrinello-Rahman barostat~\cite{parrinello1981polymorphic}~\bluec{(in standard Gromacs implementation)} with $\tau_P = 2.0~\textsf{ps}$ to keep
the pressure at ambient condition (1~Bar). Since the
change of the system size is small and slow, the pressure control
does not have an sizable effect on the
dynamics of the system.
% \bluec{The use of leap-frog scheme and Parrinello-Rahman barostat
% was not our choice, but the default one of Gromacs.}
At time $t=0$~ps, the system
has been fully equilibrated without any EF. From $t=0$ to
the warm-up time $t=t_{\textrm{warm}}$, the EF is switched on linearly, while
after $t=t_{\textrm{warm}}$, the field is kept constant in time at
$E_{\infty}$.  In this work we consider $t_{\textrm{warm}} = 10$~ps,
and $E_{\infty} = 1$~V/nm.
The direction of the field is arbitrarily chosen along the
$x$ direction.
In this paper we denote the
EF as a function of time $\vect E(t)$.
Therefore, for the case of constant EF, we have
$\vect E(t) = (E_\infty\cdot t/t_{warm},0,0)$ for $0\leq t < t_{warm}$, and 
$\vect E(t) = (E_\infty,0,0)$
for $t \geq t_{warm}$.

@@

The change of conformation in time of the alanine dipeptide is
investigated by analyzing the probability density $p(\phi,\psi,t)$ on
the Ramachandran plot at time $t$. The definition of the dihedral
angles $\phi$ and $\psi$ is given in Fig.~\ref{fig:tmp4}~(a).
The probability density of conformations of equilibrium in absence of external field is shown in Fig.~\ref{fig:tmp4}~(b), while the same quantity, resulting from the molecular relaxation to the action of $E_{\infty} = 1$~V/nm is
given in Fig.~\ref{fig:tmp4}~(c). In the plot one can identify several clusters
in which conformations are grouped.
Therefore, in order to simplify the analysis, we divide the
Ramachandran plot into 5 subregions $\{\confaa, \confab, \confba, \confbb, \confc\}$
(see caption of Fig.~\ref{fig:tmp4} for the corresponding definition).
It is worth to notice that these conformations
are found in different secondary structures of a peptide chain:
$\confaa$ and $\confab$ 
correspond to the $\alpha$-helix. $\confba$ and $\confbb$ 
correspond to the $\beta$-sheet. $\confc$ corresponds to the
left-handed $\alpha$-helix.

From Fig.~\ref{fig:tmp4}, it is evident that the probability with which a
conformation occurs changes when the external EF is applied.
The most evident case is conformation $\confc$; in fact it is {almost} not present before the external field is applied but appears in a clear way as a response to the action of the EF. Following this line of thought, we are interested to study the change of conformation under the action of EF. Specifically, for each conformational change, we will analyze the relation between the structural relaxation of the molecule and its corresponding time scale. 
The starting point is the calculation of the probability
of each conformation:
\begin{align}
  P_I(t) = \mathbb P ( (\phi_t,\psi_t) \in I), \quad  I \textrm{ being one of the five regions } \{\confaa, \confab, \confba, \confbb, \confc\},
\end{align}
Notice that the probability is time dependent, being an observable in nonequilibrium situation.
Next,
we consider the \emph{net probability flux} from conformation $J$ to $I$, defined by:
\begin{align}\nonumber
  F_{J,I}(t) & = \frac1{\Delta t}\,
  \bigg[\,
  \mathbb P \big( (\phi_{t-\Delta t},\psi_{t-\Delta t}) \in J,\, (\phi_t, \psi_t) \in I \,\big) 
  -
  \mathbb P \big( (\phi_{t-\Delta t}, \psi_{t-\Delta t}) \in I,\, (\phi_t, \psi_t) \in J \,\big)
  \bigg], \\\label{eqn:tmp7}
  & J,I \in \{\confaa, \confab, \confba, \confbb, \confc\}.
\end{align}
% {\bf Here for $\lim_{\Delta \to 0}$ is intended that in the simulation the time interval is taken large enough so that we have  enough events and thus stable numbers (see also Ref.~\cite{schuette2011markov})}.
Where $\mathbb P\big( (\phi_{t-\Delta t}, \psi_{t-\Delta t}) \in J,\, (\phi_t, \psi_t) \in I\,\big)$
is the joint probability of the alanine being in conformation $J$ at
time $(t-\Delta t)$ \emph{and} being in conformation $I$ at time $t$.
The same definition applies to
$\mathbb P \big( (\phi_{t-\Delta t}, \psi_{t-\Delta t}) \in I,\, (\phi_t, \psi_t) \in J\,\big)$.
Here the time interval $\Delta t$ should be small enough compared to the
time scale of conformational dynamics,
so that the changes in the conformation probabilities
can be treated in linear approximation.
At the same time,  $\Delta t$ should be also large enough compared with
the time step of the MD integrator, 
so that the quantities of Eq.~\eqref{eqn:tmp7} are well-defined, and can be
estimated with sufficient numerical accuracy (see also Ref.~\cite{schuette2011markov}).
\recheck{In all the numerical examples of this work, $\Delta t = 1$~ps is used}.
A positive value of $F_{J,I}(t)$ indicates that the
net flux goes from $J$ to $I$,
while a negative value indicates a net flux from $I$ to $J$.
Through $F_{J,I}(t)$, the analysis of the nonequilibrium process 
is projected onto the analysis of the probabilistic link between discretized conformations $\{\confaa, \confab, \confba, \confbb, \confc\}$.
\recheck{ The concept of net probability flux employed by us is very close to the
  concept of kinetic rate.  However, we prefer to stick to our definition which
is mathematically simpler and univocal.}
In the case of oscillatory EF, the behaviour of the net probability flux is also highly oscillatory, thus, in order to clearly identify the essential features of the molecular conformational changes, it is convenient to study the integrated net probability flux, defined as:
\begin{align}\label{eqn:tmp7a}
  Q_{J,I} (t) = \int_0^t F_{J,I}(\tau)d \tau,
\end{align}
which expresses the cumulative effects due to the action of EF on the conformations of the molecule.

@@

\begin{figure}
  \centering
  \includegraphics[]{fig.ala.new/fig-meta-more.eps}
  \caption{The time-dependent probability of conformations $P_I(t)$ 
    when a constant EF is present, where $\sum_I P_I(t) = 1, \forall t$.
    The warm-up time is $t_{warm} = 10$~ps. The red line refers to conformation $I = \confaa$,
    green to $\confab$, dark blue to $\confba$, purple to $\confbb$ and light blue
    to $\confc$. \recheck{
      The solid lines denote the results of one branching trajectory
      starting from each initial conformation, while the dashed lines denote
      the average of the four branching trajectories (different random seeds for the Langevin thermostat, see text) starting from each initial
      conformation.
      The shadowed region with each solid line denotes the
      statistical uncertainty of the result at 95\% confidence level.
    }
  }
  \label{fig:tmp5}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.3\textwidth]{fig.ala.new/fig-trans-010-flux-1.eps}
  \includegraphics[width=0.3\textwidth]{fig.ala.new/fig-trans-010-flux-2.eps}
  \includegraphics[width=0.3\textwidth]{fig.ala.new/fig-trans-010-flux-3.eps}
  \includegraphics[width=0.3\textwidth]{fig.ala.new/fig-trans-010-flux-4.eps}
  \includegraphics[width=0.3\textwidth]{fig.ala.new/fig-trans-010-flux-5.eps}
  \caption{
    The net probability flux for the constant EF case.
    The warm-up time is $t_{warm} = 10$~ps.
    The probability flux $F_{J,I}(t)$ (defined by Eqn.~\eqref{eqn:tmp7}) is reported in units of $\textrm{ps}^{-1}$.
    From top, left to right we report $I = \confaa$, $\confab$, $\confba$; bottom, left to right $\confbb$ and
    $\confc$, respectively. In each plot, the red line stands for $J=\confaa$,
    green for $J=\confab$, dark blue for $J=\confba$, purple for $J=\confbb$ and light blue
    for $J=\confc$.
    \bluec{This plot is drawn from the four-branching-trajectories case (see the text).}
    }
  \label{fig:tmp6}
\end{figure}
The time-dependent probability density of the system to stay in a certain
conformation is given in
Fig.~\ref{fig:tmp5}.
\recheck{In order to show that  our simulation results are robust with respect to the randomness introduced by the (local) Langevin thermostat, we have tested the effects of changing random seeds for the thermostat in the thermostated region. We compared simulation results obtained by starting a branching trajectory
  from each initial conformation with those obtained from four different branching trajectories from each initial conformation (each of four with
  different random seed for the Langevin thermostat in the thermostated region).
  Results are consistent within the statistical uncertainty (see Fig.\ref{fig:tmp5}).
}
The $\beta$-sheet conformations $\confba$ and $\confbb$,
which are present at the initial stage after EF is switched on,
\recheck{fade away as the system relaxes}. The
probability of the $\alpha$-helix conformation $\confaa$, grows from 0.45 to 0.69.
The $\alpha$-helix $\confab$  does not change significantly  under the action of EF.
The left-hand helix conformation $\confc$ noticeably grows from 0 to 0.22.
We plot the net probability fluxes \bluec{of the four-branching-trajectories case} in Fig.~\ref{fig:tmp6}.
Results show that for short $t$, the net fluxes are generally non-zero,
however, 
as $t$ goes to infinity,
all fluxes converge to zero. This indicates
that as the EF is switched on, the system is driven away from the initial
equilibrium (at zero EF),  starting a dynamical nonequilibrium process.
\recheck{After sufficiently long time, the system
is fully relaxed to the new stationary state, in
which it remains as long as the EF is switched on.
Of course we cannot exclude very slow nonequilibrium processes,
which cannot be captured by the duration of
our nonequilibrium simulation (1~ns), since it is
obviously that the proposed nonequilibrium approach
can only study time-dependent behaviours that are shorter than
the total time of the branching trajectories. 
}
% After a time long enough, the system
% is fully relaxed to the new equilibrium defined by the action of the constant EF,
% and stays there as long as the EF is kept.
An important point is that despite the system is finally relaxed to the new conformation,
the relaxation timescales corresponding to each conformation are rather different. 
From Fig.~\ref{fig:tmp5} and \ref{fig:tmp6}, we observe mainly
three different timescales 10~ps, 100~ps and 500~ps,
whose corresponding conformations are summarized in Tab.~\ref{tab:tmp1}.
This is a relevant result because it shows the possibility of employing an external EF as a tool to identify relevant time scales in the conformational behaviour of a molecule in solution.
